\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-07 1-intro.Rmd
all the  -> \code{CODE} as per jss  guideline
spelling

Currently, available spatio-temporal data structure in R includes: \pkg{spacetime} [@spacetime], which proposes four space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI), and trajectory (STT). The data structure it uses is based on \pkg{sp} [@sp] and \pkg{xts} [@xts], both of which has been replaced by more recent implementations. \pkg{spatstat} [@spatstat] implements a `ppp` class for point pattern data; and more recent, \pkg{stars} [@stars] implements a spatio-temporal array with the dplyr's data cube structure \pkg{cubelyr} [@cubelyr] as its backend. While these implementations either store spatial and temporal variables all in a single table, hence duplicate the spatial variables for each temporal unit; or split them into two separate tables that has the problem of manually joining, mentioned in the previously. This creates a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: the Ireland wind data from \pkg{gstat} is an classic example data that splits variables into its spatial (`wind.loc`) and temporal (`wind`) dimension; Bureau of Meteorology (BoM) provides climate observations widely applied in agriculture and ecology study. \newline

This paper describes the implementation of a new spatio-temporal data structure: \pkg{cubble}. \pkg{cubble} implements a relational data structure that uses two forms to manage the switch between spatial and temporal dimension. With this structure, users can manipulate the spatial or temporal dimension separately, while leaves the linking of two dimensions to \pkg{cubble}. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link]. \newline

The rest of the paper will be divided as follows: Section 2 presents the workflow of data manipulation in cubble. Section 3 explains how cubble deals with some advanced considerations  including data with hierarchical structure, data matching, and how cubble fits with existing static and interactive visualisation tools. Section 4 gives examples of the features introduced in the previous two sections with climate and hydrology data. An example on how cubble handle Netcdf data is also provided. Section 5 concludes the paper.

>>>>

Currently, available spatio-temporal data structure in R includes: \pkg{spacetime} [@spacetime], which proposes four space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI), and trajectory (STT). The data structure it uses is based on \pkg{sp} [@sp] and \pkg{xts} [@xts], both of which has been replaced by more recent implementations. \pkg{spatstat} [@spatstat] implements a `ppp` class for point pattern data; and more recent, \pkg{stars} [@stars] implements a spatio-temporal array with the dplyr's data cube structure \pkg{cubelyr} [@cubelyr] as its back-end. While these implementations either store spatial and temporal variables all in a single table, hence duplicate the spatial variables for each temporal unit; or split them into two separate tables that has the problem of manually joining, mentioned in the previously. This creates a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: the Ireland wind data from \pkg{gstat} is an classic example data that splits variables into its spatial (`wind.loc`) and temporal (`wind`) dimension; Bureau of Meteorology (BoM) provides climate observations widely applied in agriculture and ecology study.

This paper describes the implementation of a new spatio-temporal data structure: \pkg{cubble}. \pkg{cubble} implements a relational data structure that uses two forms to manage the switch between spatial and temporal dimension. With this structure, users can manipulate the spatial or temporal dimension separately, while leaves the linking of two dimensions to \pkg{cubble}. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

The rest of the paper will be divided as follows: Section 2 presents the workflow of data manipulation in cubble. Section 3 explains how cubble deals with some advanced considerations  including data with hierarchical structure, data matching, and how cubble fits with existing static and interactive visualisation tools. Section 4 gives examples of the features introduced in the previous two sections with climate and hydrology data. An example on how cubble handle NetCDF data is also provided. Section 5 concludes the paper.

### 2022-02-07 3-advanced.Rmd
spelling

This functionality can be seen as a matching [@stuart2010matching; @mcintosh2018using] exercise in the spatio-temporal domain to pair up nearby observatoins. It can also be considered as a form of data fusion [@castanedo2013review; @cocchi2019data], where data from multiple sources are combined together.

>>>>

This functionality can be seen as a matching [@stuart2010matching; @mcintosh2018using] exercise in the spatio-temporal domain to pair up nearby observations. It can also be considered as a form of data fusion [@castanedo2013review; @cocchi2019data], where data from multiple sources are combined together.

### 2022-02-07 4-4-netcdf.Rmd
spelling

Memory limit with Netcdf data in cubble depends on longitude grid point x latitude grid point x time grid point x number of variable. Cubble can handle slightly more than 300 x 300 (longitude x longitude) grid points for 3 variables in one year and spatial grid can be reduced to trade for longer time period and more variables. A 300 by 300 spatial grid can be a bounding box of [100, -80, 180, 0] at 0.25 degree resolution or global bounding box [-180, -90, 180, -90] at 1 degree resolution. Subsetting longitude and latitude grid is available through \code{long_range} and \code{lat_range} if the Netcdf file has finer resolution than needed.

>>>>

Memory limit with NetCDF data in cubble depends on longitude grid point x latitude grid point x time grid point x number of variable. Cubble can handle slightly more than 300 x 300 (longitude x longitude) grid points for 3 variables in one year and spatial grid can be reduced to trade for longer time period and more variables. A 300 by 300 spatial grid can be a bounding box of [100, -80, 180, 0] at 0.25 degree resolution or global bounding box [-180, -90, 180, -90] at 1 degree resolution. Subsetting longitude and latitude grid is available through \code{long_range} and \code{lat_range} if the NetCDF file has finer resolution than needed.

### 2022-02-07 5-conclusion.Rmd
spelling

This paper describes the \proglang{R} package \pkg{cubble} for manipulating and visualising spatio-temporal data. A new data structure, that builds from the \code{rowwise_df} and \code{grouped_df} class in the tidyverse ecosystem, is proposed to connect the spatial and temporal variables in the spatio-temporal data. This design frees the data analysts from spending time on organising variables of different observational units. The data structure is also flexible to the techniques and packages analysts use to analyse the data, for example, in the matching example in section 4.3, users are free to use algorithms from another pacakge to cluster stations. Further development and maintenance of the package involves responding to changes in the tidyverse packages that \pkg{cubble} imports, in particular, \pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}. Another area for further development is to extend \pkg{cubble} to other spatial objects other than points.

>>>>

This paper describes the \proglang{R} package \pkg{cubble} for manipulating and visualising spatio-temporal data. A new data structure, that builds from the \code{rowwise_df} and \code{grouped_df} class in the tidyverse ecosystem, is proposed to connect the spatial and temporal variables in the spatio-temporal data. This design frees the data analysts from spending time on organising variables of different observational units. The data structure is also flexible to the techniques and packages analysts use to analyse the data, for example, in the matching example in section 4.3, users are free to use algorithms from another package to cluster stations. Further development and maintenance of the package involves responding to changes in the tidyverse packages that \pkg{cubble} imports, in particular, \pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}. Another area for further development is to extend \pkg{cubble} to other spatial objects other than points.

### 2022-02-07 main.tex
spelling

Currently, available spatio-temporal data structure in R includes:
\pkg{spacetime} \citep{spacetime}, which proposes four space-time
layouts: Full grid (STF), sparse grid(STS), irregular (STI), and
trajectory (STT). The data structure it uses is based on \pkg{sp}
\citep{sp} and \pkg{xts} \citep{xts}, both of which has been replaced by
more recent implementations. \pkg{spatstat} \citep{spatstat} implements
a \texttt{ppp} class for point pattern data; and more recent,
\pkg{stars} \citep{stars} implements a spatio-temporal array with the
dplyr's data cube structure \pkg{cubelyr} \citep{cubelyr} as its
backend. While these implementations either store spatial and temporal
variables all in a single table, hence duplicate the spatial variables
for each temporal unit; or split them into two separate tables that has
the problem of manually joining, mentioned in the previously. This
creates a gap in the software development. The requirement for such a
tool is important given the ubiquity of spatio-temporal vector data in
the wild: the Ireland wind data from \pkg{gstat} is an classic example
data that splits variables into its spatial (\texttt{wind.loc}) and
temporal (\texttt{wind}) dimension; Bureau of Meteorology (BoM) provides
climate observations widely applied in agriculture and ecology study.

@@

The rest of the paper will be divided as follows: Section 2 presents the
workflow of data manipulation in cubble. Section 3 explains how cubble
deals with some advanced considerations including data with hierarchical
structure, data matching, and how cubble fits with existing static and
interactive visualisation tools. Section 4 gives examples of the
features introduced in the previous two sections with climate and
hydrology data. An example on how cubble handle Netcdf data is also
provided. Section 5 concludes the paper.

@@

This functionality can be seen as a matching
\citep{stuart2010matching, mcintosh2018using} exercise in the
spatio-temporal domain to pair up nearby observatoins. It can also be
considered as a form of data fusion
\citep{castanedo2013review, cocchi2019data}, where data from multiple
sources are combined together.

@@

Memory limit with Netcdf data in cubble depends on longitude grid point
x latitude grid point x time grid point x number of variable. Cubble can
handle slightly more than 300 x 300 (longitude x longitude) grid points
for 3 variables in one year and spatial grid can be reduced to trade for
longer time period and more variables. A 300 by 300 spatial grid can be
a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree resolution or
global bounding box {[}-180, -90, 180, -90{]} at 1 degree resolution.
Subsetting longitude and latitude grid is available through
\code{long_range} and \code{lat_range} if the Netcdf file has finer
resolution than needed.

>>>>

Currently, available spatio-temporal data structure in R includes:
\pkg{spacetime} \citep{spacetime}, which proposes four space-time
layouts: Full grid (STF), sparse grid(STS), irregular (STI), and
trajectory (STT). The data structure it uses is based on \pkg{sp}
\citep{sp} and \pkg{xts} \citep{xts}, both of which has been replaced by
more recent implementations. \pkg{spatstat} \citep{spatstat} implements
a \texttt{ppp} class for point pattern data; and more recent,
\pkg{stars} \citep{stars} implements a spatio-temporal array with the
dplyr's data cube structure \pkg{cubelyr} \citep{cubelyr} as its
back-end. While these implementations either store spatial and temporal
variables all in a single table, hence duplicate the spatial variables
for each temporal unit; or split them into two separate tables that has
the problem of manually joining, mentioned in the previously. This
creates a gap in the software development. The requirement for such a
tool is important given the ubiquity of spatio-temporal vector data in
the wild: the Ireland wind data from \pkg{gstat} is an classic example
data that splits variables into its spatial (\texttt{wind.loc}) and
temporal (\texttt{wind}) dimension; Bureau of Meteorology (BoM) provides
climate observations widely applied in agriculture and ecology study.

@@

The rest of the paper will be divided as follows: Section 2 presents the
workflow of data manipulation in cubble. Section 3 explains how cubble
deals with some advanced considerations including data with hierarchical
structure, data matching, and how cubble fits with existing static and
interactive visualisation tools. Section 4 gives examples of the
features introduced in the previous two sections with climate and
hydrology data. An example on how cubble handle NetCDF data is also
provided. Section 5 concludes the paper.

@@

This functionality can be seen as a matching
\citep{stuart2010matching, mcintosh2018using} exercise in the
spatio-temporal domain to pair up nearby observations. It can also be
considered as a form of data fusion
\citep{castanedo2013review, cocchi2019data}, where data from multiple
sources are combined together.

@@

Memory limit with NetCDF data in cubble depends on longitude grid point
x latitude grid point x time grid point x number of variable. Cubble can
handle slightly more than 300 x 300 (longitude x longitude) grid points
for 3 variables in one year and spatial grid can be reduced to trade for
longer time period and more variables. A 300 by 300 spatial grid can be
a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree resolution or
global bounding box {[}-180, -90, 180, -90{]} at 1 degree resolution.
Subsetting longitude and latitude grid is available through
\code{long_range} and \code{lat_range} if the NetCDF file has finer
resolution than needed.
